---
title: "Metro City School District Compensation Analysis"
author: "Analyst at TNTP"
date: "`r format(Sys.time(), '%B %d, %Y')`" # embeds the date the file was last updated
output:
  word_document:
    reference_docx: tntp-style-file.docx
    toc: no
---

This is **an R Markdown document**. Markdown allows you to make tables and figures directly from your code, rather than copying and pasting them into a Word document. It's the ultimate level of reproducability, as you'll never have to check if the numbers were copied correctly.

Hit the "Knit Word" button above to see how this document can easily turn into a simple document.

When you click the Knit button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. In other words, you can have the knitted document show your results (like a table or graph) *and* your code if you want.

Below, we have chunked off your analysis for you and given you suggestions and hints for writing the code that will create the table and graphs called for.

Ready to get started? Good! Because we want to make this look somewhat like a memo, delete all of this intro text above the first Markdown chunk (but keep the header text between the "---"s at the very top) and add your names to the header as the author.


```{r packages, include = FALSE}

# Load basic packages

library(pacman)
p_load(dplyr, tidyr, readr, stringr, ggplot2, knitr, scales)

### The official TNTP colors are:
### dark_blue = "#034772", medium_blue = "#2888BC", light_blue = "#73B7CE", green = "#699D46", orange = "#EA8936", 
### gold = "#F9C347", dark_gray = "#58595B", medium_gray = "#7D7E81", light_gray = "#C1C2C4"


```


```{r set_knitr_defaults, include = FALSE}

# This sets default options for each knitr chunk, to save typing them over and over. You can ignore for now, and override in any given chunk if desired.

opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
opts_chunk$set(dev = "png", fig.width = 9, fig.height = 5.5, dpi = 300) # control image sizes 
opts_chunk$set(size = "small", background = "white")
opts_chunk$set(highlight = TRUE, comment = NA, tidy = FALSE)
opts_chunk$set(knitr.table.format = "html")

```


```{r import, include = FALSE}

# Read in data and order the rating variable from worst to best rating.

mcsd_data <- read_csv("Module 2 Data FINAL.csv") %>%
  mutate(rating = factor(rating,
                         levels = c("Ineffective",
                                    "Minimally Effective",
                                    "Developing",
                                    "Proficient",
                                    "Highly Effective")))

```


### Question 0 (as a group): what is the distribution of ratings in MCSD?

Most teachers in MCSD are rated "Developing" (60%), with more teachers rated below "Developing" than above and very few teachers earning the extreme ratings ("Ineffective" and "Highly Effective"):

```{r q0a}

ratings_dist <- mcsd_data %>%
  group_by(rating) %>%
  summarize(count = n()) %>%
  mutate(percent = count / sum(count)) %>%
  select(rating, percent, count) # choose and reorder columns 
  
kable(ratings_dist, col.names = c("Rating", "Percent", "N"), align = "lrr") # print as a table in Word

```


```{r q0b}

ggplot(mcsd_data, aes(x = score)) +
  geom_histogram(binwidth = 2, color = "white") +
  labs(title = "MCSD 2015-16 Distribution of Ratings") +
  theme_minimal()

```



### Question 1: Confirm the problem

Developing teachers are retained at the highest rate, while Proficient and Highly Effective teachers are being retianed at rates lower than their less effective peers.

```{r q1_table}

retention_dist <- mcsd_data %>%
  group_by(rating) %>%
  summarise(percent_retained = percent(sum(retained)/n()))

kable(retention_dist, col.names = c("Rating", "Retention_Rate", "N"), align = "lrr") # print as a table in Word


# Create a summary table that shows percent retention by rating category among all teachers.
#   Hints: Because retained is coded as 1-0, the mean of this variable is the same thing as the proportion retained.
#          Start with your data and use the pipe %>% to group_by(), then summarize().
#          Finally, call kable() to display your table nicely in the knitted output document.


```



### Question 2: Explore the relationship between teacher experience and performance.

Years of experience has minimal impact on performance evaluation scores. Scores only increase slightly for every decade of additional teaching experience. Need to create object using regression model.

```{r q2}
experience_and_performance <- lm(mcsd_data$score ~ mcsd_data$experience)

ggplot(mcsd_data, aes(x = experience, y = score)) +
  geom_point() +
  geom_jitter(alpha = .3) +
  geom_smooth(method = "lm", formula = y~x) +
  labs(x = "Years of Teaching Experience", y = "Performance Evaluation Score") +
  theme_light()

# Create a figure showing the relationship between years of experience and
# evaluation score. Consider geom_point, geom_jitter, and/or geom_smooth. If you
# have problems with overplotting, lower the alpha value within geom_point or
# geom_jitter, or plot only the mean value for each experience year.


```



### Question 3: Document the (lack of) relationship between highest degree held and teacher performance.

Add your summary sentences here.

```{r q3}


ggplot(mcsd_data, aes(x = degree, y = score)) +
  geom_boxplot() +
  labs(x = "Highest Degree Earned", y = "Performance Evaluation Score")


# Create a figure showing the relationship between degree and evaluation score.
# With one categorical variable and one continuous, consider geom_boxplot,
# geom_violin, or geom_jitter, or display only the means with geom_bar.


```



### Question 4: Explore the connection between compensation and performance.

Salary varies greatly within each of the 5 rating groups. There are many instances of teachers making similar salaries with vastly disparate scores. 

```{r q4}

ggplot(mcsd_data, aes(x = salary, y = score, col = rating)) +
  geom_point() +
  geom_smooth() +
  labs(x = "Salary", y = "Performance Evaluation Score", legend = "Evaluation Rating") +
  scale_color_manual(values= c("#034772", "#2888BC", "#73B7CE", "#699D46", "#EA8936"))

# Create a plot comparing performance and compensation.  Consider geom_point,
# geom_jitter, and geom_smooth.
# Consider adding a 3rd variable as a color, for example experience category,
# or high-performers vs. low performers; or high-paid vs. low-paid.


```



### Question 5: Contrast the differences between early and late career teachers in performance vs. in pay.

More experience teachers, on average, make more than less experienced teachers, regardless of performance ratings. While slightly more teachers rated "ineffective" or "minimally effective" are novice, both veteran and novice teacher ratings are similarly distributed across ratings. Compensation has a much closer relationship with pay than evaluation score does. 

```{r q5_bonus}

# Create the grouping variable using a cutoff number of year of experience.


career_experience <- mcsd_data %>%
  mutate(experience, teaching_experience = ifelse(experience <= 10, "Early Career", "Late Career"))


ggplot(career_experience, aes(score, col=teaching_experience)) + 
  geom_density() + 
  labs(title = "MCSD:Performance based on teaching experience") +
  scale_color_manual(values= c("#034772", "#2888BC", "#73B7CE", "#699D46", "#EA8936"))

ggplot(career_experience, aes(salary, col=teaching_experience)) + 
  geom_density() + 
  labs(title = "MCSD:Compensation based on teaching experience") +
  scale_color_manual(values= c("#034772", "#2888BC", "#73B7CE", "#699D46", "#EA8936")) 

ggplot(career_experience, aes(rating, salary, color=teaching_experience)) + 
  geom_jitter() + 
  labs(title = "MCSD:Compensation and Performance based on teaching experience") +
  scale_color_manual(values= c("#034772", "#73B7CE"))

ggplot(career_experience, aes(score, salary, color=teaching_experience)) + 
  geom_jitter() + 
  labs(title = "MCSD:Compensation and Performance based on teaching experience") +
  scale_color_manual(values= c("#034772", "#73B7CE"))



# Compare the distribution of salaries between the two groups...
# Consider geom_histogram or geom_density


# ... and then compare the performance scores of new vs. veteran teachers using the same geom:


```
